Sample-efficient Reinforcement Learning \\ with Implicitly Quantized Representations

### 2024-02-01 paper/main.tex
adding to method

\begin{abstract}
  In reinforcement learning (RL) for continuous control, we propose an
  efficient representation learning method using only a latent-state
  consistency loss. Our approach employs an encoder and a transition
  model to map observations to latent states and predict future
  states, respectively. We achieve high performance and prevent
  representation collapse by normalizing, quantizing, and bounding the
  latent representation. Our method is straightforward, compatible
  with any model-free RL algorithm, and demonstrates state-of-the-art
  performance in continuous control benchmarks from DeepMind Control
  Suite.
\end{abstract}

%% \begin{abstract}
%%   Learning representations for reinforcement learning (RL) has shown
%%   much promise for continuous control. Representations not only can
%%   handle complex observation spaces but also improve sample
%%   efficiency. A particularly promising approach is to use a
%%   self-supervised latent-state consistency loss with observation to
%%   latent state mapping, together with next latent state, reward, and
%%   value predictions. Reward and value predictions prevent
%%   representation collapse at the cost of learning a task-specific
%%   representation.  We show how to use only the latent-state
%%   consistency loss to {\em (i)} learn a task-agnostic representation
%%   and {\em (ii)} prevent representation collapse by normalizing and
%%   bounding the latent representation. In experiments, our approach
%%   which be combined with any model-free RL algorithm obtains
%%   state-of-the-art results in continuous control benchmarks from
%%   DeepMind Control Suite.
%% \end{abstract}

%% \begin{abstract}
%%   % Deep reinforcement learning (RL) has shown much promise for solving real-world continuous control problems.
%%   % Whilst deep reinforcement learning (RL) is a competitive approach for solving real-world continuous control problems, it
%%   % is limited by its sample-efficiency.
%%   % Recent works have shown that learning representations can improve sample efficiency in RL, whilst also
%%   % handling complex/high-dimensional observation spaces.
%%   Learning representations for reinforcement learning (RL) has shown much promise for continuous control.
%%   Not only can it handle complex observation spaces but it can also improve sample efficiency.
%%   % Recent works have shown that learning representations can improve sample efficiency in RL.
%%   A particularly promising approach is to use a self-supervised loss, known as the latent-state consistency loss.
%%   It requires an encoder which maps observations to latent states and a transition model which predicts
%%   next latent states.
%%   It then minimizes the distance between the next state predicted by the transition model and
%%   the next state predicted by the encoder.
%%   Typically it is combined with other loss terms such as reward or value prediction.
%%   This prevents representation collapse at the cost of learning a task-specific representation.
%%   In this work, we show that a careful consideration of the latent space enables us to {\em (i)} learn
%%   a task-agnostic representation and {\em (ii)} prevent representation collapse, whilst using only the latent-state
%%   consistency loss.
%%   % the leverage the latent-state
%%   % Typically these approaches have unbounded latent spaces and are combined with auxillary loss terms such as minimizing the
%%   % error of reward or value predictions in the latent space.
%%   % of an encoder which maps observations to latent states and a transition model which predicts
%%   % next latent states given a latent state and an action.
%%   % However, most of th
%%   We empirically show that our method prevents representation collapse as it maintains matrix rank.
%%   Our representation learning approach is {\em (i)} simple, {\em (ii)} can be combined with any model-free RL algorithm
%%   and, {\em (iii)} obtains state-of-the-art results in continuous control benchmarks from DeepMind Control Suite.
%%   % Moreover, it kk
%% % This document provides a basic paper template and submission guidelines.
%% % Abstracts must be a single paragraph, ideally between 4--6 sentences long.
%% % Gross violations will trigger corrections at the camera-ready phase.
%% \end{abstract}

% Given such a latent space,
% Whilst these latent states are often lower dimensional this is not always
Many approaches have been proposed to improve sample efficiency in deep RL.
As \citet{laskinCURLContrastiveUnsupervised2020} highlighted, these can generally be divided into two streams of research:
{\em (i)} auxillary tasks on the agent's observations and {\em (ii)} world models that predict the future
\citep{haRecurrentWorldModels2018,hafnerLearning2019,hansenTemporalDifferenceLearning2022}.
In this work, we are interested in the first class of methods, which use auxillary tasks to improve sample efficiency.
In particular, methods for compressing observations into latent states.%, aka representation learning.

Learning representations for RL has been investigated for decades
\citep{abelOptimalBehaviorApproximate2016,mannorDynamicAbstractionReinforcement2004,liUnifiedTheoryState2006,andreStateAbstractionProgrammable2002,deardenAbstractionApproximateDecisiontheoretic1997,singhReinforcementLearningSoft1994,higginsDefinitionDisentangledRepresentations2018,vanhoofStableReinforcementLearning2016,watterEmbedControlLocally2015,ghoshRepresentationsStableOffPolicy2020}.
However, these approaches are usually limited to simple environments.
Some approaches used unsupervised learning (\eg VAE \citep{kingmaAutoEncoding2014} with reconstruction loss) to acquire representations
\citep{finnDeepSpatialAutoencoders2016,higginsDARLAImprovingZeroShot2017,langeAutonomousReinforcementLearning2012,watterEmbedControlLocally2015}.
More recently, self-supervised learning (SSL) approaches (which do not reconstruct observations)
attempt to learn good features without labels \cite{anandUnsupervisedStateRepresentation2019}.
Alternative approaches leverage contrastive losses to learn representations \cite{laskinCURLContrastiveUnsupervised2020}.

@@

In this paper, we propose a simple representation learning technique that can be combined with any model-free RL method
(we use TD3 \citep{fujimotoAddressingFunctionApproximation2018}).
See \cref{fig:diagram-self-predictive} for an overview of our (extremely simple) representation learning method.
It is based soley on the latent-state consistency loss, \ie a commonly use self-supervised loss for RL.
In contrast to prior works, we normalize our latent representation using the technique from Finite Scalar Quantization
\citep{mentzerFiniteScalarQuantization2023}.
As a result, our latent space is bounded and associated with an \textit{implicit} codebook, whose size we can control.
% As a result, our latent space is associated with an \textit{implicit} codebook, whose size we can control.
Importantly, our method {\em (i)} prevents representation collapse (as it learns a high-rank representation),
{\em (ii)} demonstrates state-of-the-art sample efficiency in continuous control benchmarks,
{\em (iii)} is simple to implement, and {\em (iii)} learns a task-agnostic
representation which could be helpful in downstream tasks.

@@

\our has four parameterized components.
\begin{align}
&\text{Encoder: } & z_{t} &= e_{\theta} (s_{t}) \label{eq:encoder} \\
&\text{Dynamics: } & \hat{z}_{t+1} &= d_{\phi} (z_{t}, a_{t}) \label{eq:transition} \\
&\text{Value: } & q_{t} &= Q_{\psi} (z_{t}, a_{t}) \label{eq:value} \\
&\text{Policy: } & a_{t} &\sim \pi_{\eta} (z_{t}) . \label{eq:policy}
\end{align}
Central to our method is the latent state $z_{t}$

>>>>

\begin{abstract}
  % Arno's proposal
  In reinforcement learning (RL) for continuous control, we propose  an efficient representation learning method using a latent-state consistency loss. This approach employs an encoder and a transition model to map observations to latent states and predict future states, respectively. By minimizing discrepancies between these states, we achieve task-agnostic representations and prevent representation collapse using only the consistency loss. Our method is straightforward, compatible with any model-free RL algorithm, and demonstrates state-of-the-art performance in continuous control benchmarks from DeepMind Control Suite.\looseness-1

  % Deep reinforcement learning (RL) has shown much promise for solving real-world continuous control problems.
  % Whilst deep reinforcement learning (RL) is a competitive approach for solving real-world continuous control problems, it
  % is limited by its sample-efficiency.
  % Recent works have shown that learning representations can improve sample efficiency in RL, whilst also
  % handling complex/high-dimensional observation spaces.
  Learning representations for reinforcement learning (RL) has shown much promise for continuous control.
  Not only can it handle complex observation spaces but it can also improve sample efficiency.
  A particularly promising approach is to use a self-supervised loss, known as the latent-state consistency loss.
  It requires an encoder which maps observations to latent states and a transition model which predicts
  next latent states.
  It then minimizes the distance between the next state predicted by the transition model and
  the next state predicted by the encoder.
  Typically it is combined with other loss terms such as reward or value prediction.
  This prevents representation collapse at the cost of learning a task-specific representation.
  In this work, we show that a careful consideration of the latent space enables us to {\em (i)} learn
  a task-agnostic representation and {\em (ii)} prevent representation collapse, whilst using only the latent-state
  consistency loss.
  We empirically show that our method prevents representation collapse as it maintains matrix rank.
  Our representation learning approach is {\em (i)} simple, {\em (ii)} can be combined with any model-free RL algorithm
  and, {\em (iii)} obtains state-of-the-art results in continuous control benchmarks from DeepMind Control Suite.
  % Moreover, it kk
% This document provides a basic paper template and submission guidelines.
% Abstracts must be a single paragraph, ideally between 4--6 sentences long.
% Gross violations will trigger corrections at the camera-ready phase.
\end{abstract}

@@

In this paper, we propose a simple representation learning technique that can be combined with any model-free RL method
(we use TD3 \citep{fujimotoAddressingFunctionApproximation2018}).
See \cref{fig:diagram-self-predictive} for an overview of our (extremely simple) representation learning method.
It is based soley on the latent-state consistency loss, \ie a commonly use self-supervised loss for RL.
In contrast to prior works, we normalize our latent representation using the technique from Finite Scalar Quantization
\citep{mentzerFiniteScalarQuantization2023}.
As a result, our latent space is bounded and associated with an \textit{implicit} codebook, whose size we can control.
% As a result, our latent space is associated with an \textit{implicit} codebook, whose size we can control.
Importantly, our method {\em (i)} alleviates representation collapse,
{\em (ii)} demonstrates state-of-the-art sample efficiency in continuous control benchmarks,
{\em (iii)} is simple to implement, and {\em (iv)} learns a task-agnostic
representation which could be helpful in downstream tasks.

@@

\our has four parameterized components.
\begin{align}
&\text{Encoder: } & z_{t} &= e_{\theta} (s_{t}) \label{eq:encoder} \\
&\text{Dynamics: } & \hat{z}_{t+1} &= d_{\phi} (z_{t}, a_{t}) \label{eq:transition} \\
&\text{Value: } & q_{t} &= Q_{\psi} (z_{t}, a_{t}) \label{eq:value} \\
&\text{Policy: } & a_{t} &\sim \pi_{\eta} (z_{t}) . \label{eq:policy}
\end{align}
The encoder $e_{\theta}$  and latent-space dynamics model $d_{\phi}$ are responsible for representation learning.
The encoder $e_{\theta}$ maps observations $o_{t}$ to latent states $z_{t}$ and is responsible for learning a representation
which can aid RL.
The latent-space dynamics model $d_{\phi}$ predicts next latent states $z_{t+1}$ given a latent state $z_{t}$ and an action $a_{t}$.
It's sole purpose is to aid representation learning by making the latent states temporally consistent.
Note that we do not use it for model-based RL.
Once we have the representation learned by our encoder, we map all observations to the latent space and perform model-free RL
in this latent space.
Throughout this paper, we use Twin Delayed Deep Deterministic Policy Gradient
(TD3), \citep{fujimotoAddressingFunctionApproximation2018}).
It consists of two state-action value functions $\{Q_{\psi_{1}},Q_{\psi_{2}} \}$, known as critics, and a deterministic
actor $\pi_{\eta}$.
Following prior works \citep{yaratsMasteringVisualContinuous2021} we use a linear exploration noise schedule
which decays from $1$ to $0.1$ during training.
